---
title: "Welcome to the 2023 R workshop!"
author: "A hopefully more-or-less painless introduction to R"
date: "UNR, Fall 2023"
output:
  html_document:
    theme: journal
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

 * **When?** Saturday Sep 16, 2023
 * **Where?** UNR (WRB 2030)  
 * **What to prepare?** Everyone should bring their own laptop, with R and RStudio installed prior to starting the workshop (see [links](Links.html)). 
 
## Facilitator

[Kevin Shoemaker](https://www.unr.edu/nres/people/shoemaker-kevin)		 
Department of Natural Resources and Environmental Science       
Office: Fleischmann Agriculture 220E					
Email: [kevinshoemaker@unr.edu](kevinshoemaker@unr.edu)

[Carlos Ramirez-Reyes](https://library.unr.edu/ramirez-reyes-carlos)		 
UNR Data Services Coordinator       
DeLaMare Science and Engineering Library					
Email: [cramirezreyes@unr.edu](cramirezreyes@unr.edu)


## Contributors

[Perry Williams](http://www.perrywilliams.us/)  
Department of Natural Resources and Environmental Science   
Office: 240 Fleischmann  
Email: perryw(at)unr(dot)edu

[Mitchell Gritts](https://mitchellgritts.com)     
Vibrant Planet          
Email: mitchell(at)gritts(dot)me

[Christine Albano](http://www.csp-inc.org/about-us/core-science-staff/albano-christine/)     
Lead Scientist at Conservation Science Partners and doctoral student at DRI    
Email: christine(at)csp-inc(dot)org    

[Stephanie Freund](http://newinghamlab.weebly.com/lab-members.html)      
Office: Knudtsen Resource Center 119
Department of Natural Resources and Environmental Science / USDA Agricultural Research Service
Email: smfreu(at)gmail(dot)com

[Jessi Brown](http://jessilbrown.org/Welcome.html)    
Department of Biology     
Office: Fleischmann Agriculture 244       
Email: jlbrown(at)unr(dot)edu    

[Jonathan Greenberg](https://naes.unr.edu/gears/)    
Department of Natural Resources and Environmental Science   
Office: 241 Fleischmann Ag    
Email: jgreenberg(at)unr(dot)edu     

[Ken Nussear](http://www.nussear.com/)  
Department of Geography      
Office: Mackay Science Room 223      
Email: knussear(at)unr(dot)edu        

[Paul Hurtado](http://pauljhurtado.com/)  
Department of Mathematics and Statistics     
Office: Davidson Math and Science Center 220  
Email: [phurtado@unr.edu](phurtado@unr.edu)    


## About

The statistical programming software 'R' is one of the fundamental tools for modern data exploration and analysis, and a basic ability to use R (for data processing, statistical analysis, simulation modeling and production of high-quality figures) will make upcoming classes, research, and graduate school less intimidating and more productive (and fun!).

This workshop consists of a series of short(ish) submodules, each of which covers a particular aspect of working with data (e.g., read and writing data to/from files, making exploratory plots). Each submodule will consist of a set of lectures and demos (worked examples), followed by hands-on activities. The main goal of this workshop will be to ensure participants have enough proficiency and confidence with data operations and programming in R to engage in productive, self-directed learning and problem-solving. 

The workshop is primarily intended for students with little prior experience with R. The first modules will focus on R syntax, data management (loading data, writing to file), data visualizations, and data "wrangling". The second set of modules will focus on R packages (loading, getting help), programming operations (loops, functions, debugging etc.), statistics, and spatial analyses.   

All code will be available as scripts that you can download from this website (at the top of each module page on this website) and load in RStudio. That way you won't need to constantly copy and paste from the website!  

## Before we get started...

Before we dig in and get started with the modules, you should have installed R and RStudio. Here are some links to help you get started:

[Download and install R](https://cran.r-project.org/)          
[Download and install RStudio](https://www.rstudio.com/products/rstudio/download/)    (use free version!)         


Okay, now we're ready to go!


[--go to first module--](module1_1.html)


```{r echo=FALSE}
rmd2rscript <- function(infile="module1_1.Rmd"){    # function for converting markdown to scripts
  outfile1 <- gsub(".Rmd",".R",infile)
  outfile2 <- gsub(".Rmd",".txt",infile)
  close( file( outfile1, open="w" ) )   # clear output file
  close( file( outfile2, open="w" ) )   # clear output file
  con1 <- file(infile,open="r")
  con2 <- file(outfile1,"w")
  con3 <- file(outfile2,"w")
  stringToFind <- "```{r*"
  isrblock <- FALSE
  count=0
  
  while(length(input <- readLines(con1, n=1)) > 0){   # while there are still lines to be read
    isrblock <- grepl(input, pattern = stringToFind, perl = TRUE)   # is it the start of an R block?
    if(isrblock){
      while(!grepl(newline<-readLines(con1, n=1),pattern="```",perl=TRUE)){
        if(count>1){
          write(newline,file=con2,append=TRUE)
          write(newline,file=con3,append=TRUE)
        }
        count=count+1
      }
      isrblock=FALSE
    }
  }
  closeAllConnections()
}

rmd2rscript2 <- function(infile="LECTURE2.Rmd"){    # function for converting markdown to scripts
  outfile <- gsub(".Rmd",".R",infile)
  close( file( outfile, open="w" ) )   # clear output file
  con1 <- file(infile,open="r")
  con2 <- file(outfile,"w")
  stringToFind <- "```{r*"
  stringToFind2 <- "echo"
  isrblock <- FALSE
  #count=0
  blocknum=0
  
  while(length(input <- readLines(con1, n=1)) > 0){   # while there are still lines to be read
    isrblock <- grepl(input, pattern = stringToFind, perl = TRUE)   # is it the start of an R block?
    showit <- !grepl(input, pattern = stringToFind2, perl = TRUE)   # is it hidden (echo=FALSE)
    if(isrblock){
      blocknum=blocknum+1
      while(!grepl(newline<-readLines(con1, n=1),pattern="```",perl=TRUE)){
        if((blocknum>1)&((showit)|(blocknum==2))) write(newline,file=con2,append=TRUE)
        #count=count+1
      }
      isrblock=FALSE
    }
  }
  closeAllConnections()
}

```


```{r echo=FALSE, eval=FALSE}

rmd2rscript2("module1_1.Rmd")
rmd2rscript2("module1_2.Rmd")
rmd2rscript2("module1_3.Rmd")
rmd2rscript2("module1_4.Rmd")
rmd2rscript2("module2_1.Rmd")
rmd2rscript2("module2_2.Rmd")
rmd2rscript2("module2_3.Rmd")
rmd2rscript2("module2_4.Rmd")
rmd2rscript2("module2_5.Rmd")
#rmd2rscript2("module2_6.Rmd")

## TODO: make sure there are pdf copies of all the modules... 

knitr::opts_chunk$set(echo = TRUE, message = FALSE)
rmarkdown::render("index.Rmd",rmarkdown::pdf_document())
rmarkdown::render("module1_1.Rmd",rmarkdown::pdf_document())
rmarkdown::render("module1_2.Rmd",rmarkdown::pdf_document())
rmarkdown::render("module1_3.Rmd",rmarkdown::pdf_document())
rmarkdown::render("module1_4.Rmd",rmarkdown::pdf_document())  # latex_engine="xelatex"

rmarkdown::render("module2_1.Rmd",rmarkdown::pdf_document())
rmarkdown::render("module2_2.Rmd",rmarkdown::pdf_document())
rmarkdown::render("module2_3.Rmd",rmarkdown::pdf_document())
rmarkdown::render("module2_4.Rmd",rmarkdown::pdf_document())
rmarkdown::render("module2_5.Rmd",rmarkdown::pdf_document())


```




